---
title: "Yearly AIRSIS Data"
author: "Mazama Science"
date: "`r Sys.Date()`"
output:
  html_document: default
  html_notebook: default
---

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=5)
```

# Setup

Start by loading the PWFSLSmoke package and loading the data. 

```{r Setup, message=FALSE}
suppressPackageStartupMessages({
  library(PWFSLSmoke)
  library(ggplot2)
})

setSpatialDataDir("~/Data/Spatial")

logger.setLevel(ERROR)
load("~/Data/airsis/airsis_2015.RData")
load("~/Data/airsis/airsis_2016.RData")
load("~/Data/airsis/airsis_2017.RData")
```

# Orientation

Create some maps and plots to get an idea about what the data looks like. 

```{r mapping}
monitorLeaflet(airsis_2015)
monitorLeaflet(airsis_2016)
monitorLeaflet(airsis_2017)
```

```{r}
days2015 <- sum(!is.na(airsis_2015$data[-1]))/24
days2015
days2016 <- sum(!is.na(airsis_2016$data[-1]))/24
days2016
days2017 <- sum(!is.na(airsis_2017$data[-1]))/24
days2017
deployments2015 <- ncol(airsis_2015$data) - 1
deployments2015
deployments2016 <- ncol(airsis_2016$data) - 1
deployments2016
deployments2017 <- ncol(airsis_2017$data) - 1
deployments2017
```

There were `r deployments2015` different deployments in 2015, `r deployments2016` deployments in 2016, and `r deployments2017` deployments in 2017. Monitors were deployed for a combined total of `r days2015` days in 2015, `r days2016` days in 2016, and `r days2017` days in 2017. This means that each deployment lasted an average of `r days2015/deployments2015` days in 2015, `r days2016/deployments2016` days in 2016, and `r days2017/deployments2017` days in 2017.  

Let's see how they were distributed through time:

```{r deployment_boxplot, fig.height=7, cache = TRUE}
# Create dataframes for each year with a column for the time, and another for the 
# number of valid data points that hour
counts2015 <-  data.frame(time = airsis_2015$data$datetime, 
                          count = apply(airsis_2015$data[-1], 1, function(x) sum(!is.na(x))))
counts2016 <-  data.frame(time = airsis_2016$data$datetime, 
                          count = apply(airsis_2016$data[-1], 1, function(x) sum(!is.na(x))))
counts2017 <-  data.frame(time = airsis_2017$data$datetime, 
                          count = apply(airsis_2017$data[-1], 1, function(x) sum(!is.na(x))))

# Create plots of the number of valid data points per hour
plot1 <- ggplot(counts2015, aes(time, count))+
  geom_area(col = "gray20")+
  geom_line(col = "gray20")+
  labs(title='Number of valid data points by hour', subtitle='2015')+
  ylim(0,35)
plot2 <- ggplot(counts2016, aes(time, count))+
  geom_area(col = "gray20")+
  geom_line(col = "gray20")+
  labs(subtitle='2016')+
  ylim(0,35)
plot3 <- ggplot(counts2017, aes(time, count))+
  geom_area(col = "gray20")+
  geom_line(col = "gray20")+
  labs(subtitle='2017')+
  ylim(0,35)

# Arrange all plots together
gridExtra::grid.arrange(plot1, plot2, plot3, nrow = 3)
```

Using monitor_combine, we can combine all three years into one monitor object and then make the previous plot on a continuous timeline.

```{r long_deployment_boxplot, cache = TRUE}
# Combine the monitors
airsis <- monitor_combine(list(airsis_2015, airsis_2016))
airsis <- monitor_combine(list(airsis, airsis_2017))

# Get the deployment counts
counts <- data.frame(time = airsis$data$datetime, 
                          count = apply(airsis$data[-1], 1, function(x) sum(!is.na(x))))

# Plot the whole thing
ggplot(counts, aes(time, count))+
  geom_area(col = "gray20")+
  geom_line(col = "gray20")+
  labs(title = "Number of valid data points by hour")

```

There are also some ways to look at summaries of the PM2.5 values throughout the months. Let's take the example of 2015 values

```{r pm25_summary, cache = TRUE, warning=FALSE, message=FALSE}
dailyMeans <- monitor_dailyStatistic(airsis_2015, FUN = get("mean"), na.rm = TRUE)
superMean <- monitor_collapse(dailyMeans, FUN = get("mean"))

dailyMax <- monitor_dailyStatistic(airsis_2015, FUN = get("max"), na.rm = TRUE)
superMax <- monitor_collapse(dailyMax, FUN = get("max"))

dailyMin <- monitor_dailyStatistic(airsis_2015, FUN = get("min"),  na.rm = TRUE)
superMin <- monitor_collapse(airsis_2015, FUN = get("min"))

dailyMedian <- monitor_dailyStatistic(airsis_2015, FUN = get("median"), na.rm = TRUE)
superMedian <- monitor_collapse(airsis_2015, FUN = get("median"))

monitorPlot_timeseries(superMax, type = "l", lwd = 1, col = adjustcolor('royalblue', alpha.f = .5))
monitorPlot_timeseries(superMin, type = "l", lwd = 1, col = adjustcolor('purple4', alpha.f = .5), add = TRUE)
monitorPlot_timeseries(superMean, type = "l", lwd = 1, col = adjustcolor('seagreen', alpha.f = .5), add = TRUE)
monitorPlot_timeseries(superMedian, type = "l", lwd = 1, col = adjustcolor('orangered4', alpha.f = .5), add = TRUE)
legend("topleft", c("max", "min", "median", "mean"), 
       col = c(adjustcolor('royalblue', alpha.f = .5), 
               adjustcolor('purple4', alpha.f = .5),
               adjustcolor('orangered4', alpha.f = .5),
               adjustcolor('seagreen', alpha.f = .5)),
       lty = c(1,1,1,1), lwd = c(1.5, 1.5, 1.5, 1.5))
title("Aggregated values from PM2.5 monitors, 2015")
```
  

Now that we've seen when the data was, let's look at where it is. There are several different options for plotting monitors from a ws_monitor object. monitorEsriMap() will plot points for each monitor with AQI colors indicating the maximum PM2.5 value using a map from ESRI, and is very similar to monitorGoogleMap except that it uses R's base plotting capabilities so it is easier to add points or polygons on top. The other option is monitorMap() which plots the points over polygons instead of a map image. 
  
```{r monthly_maps, fig.width = 10, fig.height=37.5, cache = TRUE}
# Load a basemap from ESRI
basemap <- esriMap_getMap(bboxString = '-124.68,30.19,-106.15,49.70')

# Set the layout so all maps wil be plotted in one figure
layout(matrix(1:(13*4), 13, 4, byrow = FALSE), 
       widths=c(.8,4,4,4), heights = c(1,rep(4, 13)))
par(mar = c(0,0,0,0))
plot.new()

# Write the month names
for( month in 1:12 ){
  text <- month.name[month]
  plot(0,0,col="transparent", axes = F)
  text(0,0,text, srt = 90, cex = 1.5)
}

# Plot a map of monitors for every month in each year
for ( year in c(2015, 2016, 2017) ){
  
  # Add a year label
  plot(0,0,col="transparent", axes = F)
  text(0,0, as.character(year), cex = 2)
  
  # Add a plot for each month
  for ( month in 1:12 ) {
    
    # Create the tlim
    if ( month < 12 ) {
      tlim <- paste0(as.character(year), stringr::str_pad(as.character(c(month, month+1)), 2, side = 'left', "0"), '01')
    } else {
      tlim <- paste0(as.character(year), c("120100", "123123"))
    }
    
    # Get subset the monitor object (including all years) for desired tlim
    monitor <- monitor_subset(airsis, tlim = tlim)
    
    # Try to plot the map
    # If there is an error, plot a map with no points
    result <- try (
      monitorEsriMap(monitor, mapRaster = basemap, cex = 2.5)
    )
    if ("try-error" %in% class(result)) {
      esriMap_plotOnStaticMap(basemap)
    }
    box("figure", col = 'gray50')
  }
}
```

# Northern California Fires

Northern California tends to be prone to wildfires. Let's take a look at how air quality in Northen California in each of the three years to see what stories that might be able to tell us about the fire season in 2015, 2016, and 2017 and how they changed. We'll start by just looking at monitors in Siskiyou County. 

```{r subset_monitors}
# Subset objects to include only monitors in Siskiyou county
sisk2015 <- monitor_subsetBy(airsis_2015, countyName=='Siskiyou')
sisk2016 <- monitor_subsetBy(airsis_2016, countyName=='Siskiyou')
sisk2017 <- monitor_subsetBy(airsis_2017, countyName=='Siskiyou')
```

```{r plot monitors, fig.width=9, fig.height=3}
# Set the layout so all maps will print in one figure
layout(matrix(c(1,2,3,4), 1, byrow = TRUE), widths=c(1,1)) 
par(mar = c(0,0,4,0))

# Add maps of monitors for each year
monitorMap(sisk2015, ylim = c(39.9, 42.1239), xlim = c(-124.95, -120.21), cex = 2)
box('plot')
title("2015", cex.main = 3)
monitorMap(sisk2016, ylim = c(39.9, 42.1239), xlim = c(-124.95, -120.21), cex = 2)
box('plot')
title("2016", cex.main = 3)
monitorMap(sisk2017, ylim = c(39.9, 42.1239), xlim = c(-124.95, -120.21), cex = 2)
box('plot')
title("2017", cex.main = 3)

# Add an AQI legend plot
plot.new()
addAQILegend('topleft', pt.cex = 2)
```

Now, let's take a look at the pm2.5 data from each year. 

```{r gnatplots, fig.height = 9}
# Set the layout so all plots will in one figure
layout(matrix(c(1,2,3), 3, 1), heights = c(1,1,1))
par(xaxt = "n", mar = c(0,4.1,4,3), xpd = NA)

# Add timeseries plots for each year
monitorPlot_timeseries(sisk2015, style='gnats', xlab = "", localTime = FALSE, ylim = c(0,800))
mtext("2015", 4, padj = 1)
par(xaxt = 's')
axis.POSIXct(1, sisk2015$data$datetime, labels = FALSE)
par(xaxt = 'n')
title("All pm2.5 readings from monitors in Siskiyou County")
par(mar=c(2,4.1,2,3))
monitorPlot_timeseries(sisk2016, style='gnats', xlab = "", localTime = FALSE, ylim = c(0,800))
mtext("2016", 4, padj = 1)
par(xaxt = 's')
axis.POSIXct(1, sisk2016$data$datetime, labels = FALSE)
par(mar=c(4,4.1,0,3))
monitorPlot_timeseries(sisk2017, style='gnats', xlab = "", localTime = FALSE, ylim = c(0,800))
mtext("2017", 4, padj = 1)
```

Just from looking at these plots, we can say some things about the fire season in Northern California. There is little smoke before July, and PM2.5 levels tend to peak around August or September, and then air quality returns to normal around the beginning of November. PM2.5 levels to not follow a smooth curve. During high-smoke season, PM2.5 levels tend to stay above a baseline, and spike briefly before returning to near the baseline. 

We can see that 2015 and 2017 had more smoke, and for a longer time, than 2016. Using this as a proxy for fire intensity and duration, we can say that 2015 and 2017 had longer fire seasons than 2016. It appears that there were two distinct smoke events in 2016, the first much more intense than the second. Upon closer examination, however, we see that there is no data from 2016 between September 23 and October 5. 

```{r missingData}
# Create a timeseries plot with a time limit surrounding missing values
monitorPlot_timeseries(sisk2016, tlim = c(20160920, 20161010), type = "l")
```

Let's look a little closer at the smokiest part of the summer for the three years: July to October.

```{r dygraphs}
# Create dygraphs for the summer of each year
monitorDygraph(sisk2015, tlim = c(20150701, 20151001), title = "2015")
monitorDygraph(sisk2016, tlim = c(20160701, 20161001), title = "2016")
monitorDygraph(sisk2017, tlim = c(20170701, 20171001), title = "2017")
```

Zooming in closer, a periodic pattern emerges between the dramatic smoke events. 

```{r Yreka_Dygraph}
# Plot zoomed-in dygraph
monitorDygraph(title = "2017", monitor_subset(sisk2017), tlim = c(20170708, 20170720))
```

Let's see how PM2.5 levels change throughout the day for one monitor. This plot shows the hourly PM2.5 values from every day between July 8 - July 20 overlaid on top of each other with the mean in a darker color. 

```{r monitor_spaghetti}
# pull out data for one monitor 
yreka <- monitor_subset(sisk2017, monitorID = "lon_.122.634_lat_41.727_apcd.1014")

# Plot the monitor location
monitorLeaflet(yreka)

# Create spaghetti plot of hourly values over the week
monitorPlot_timeOfDaySpaghetti(monitor_subset(yreka, tlim = c(20170708, 20170720)), title = "Yreka Jul 8 - Jul 20 2017")
addAQILegend()
```

Air quality in this location tends to peak around 9am, and disappate throughout the day. It is common to see similar patterns from day-to-day at one location. This might be due to weather patterns, smoke settling in a valley overnight, or human activites. In this case, Yreka is close to a major highway so one possibility could be that the spike around 9am corresponds to heavy traffic, and the PM2.5 levels lower as traffic slows down and wind picks up.
  
Later in the season, smoke really starts picking up in August. Let's take a closer look at the week of August 21 - 30, once again looking at the Yreka monitor.

```{r daily_barplot}
# Pull out the interesting week
yrekaWeek <- monitor_subset(yreka, tlim = c(20170821, 20170830))
monitorPlot_rollingMean(yrekaWeek)
addAQILegend()
```

The rolling mean plots plots all the points overlaid with a rolling mean line. The rolling mean helps smooth the data to make it easier to see trends.

Another built-in function for smoothing is NowCast. Here, NowCast values are plotted over the rolling mean to show the difference between NowCast and rolling mean. NowCast values will appear to trail the observed data by an hour or so. One use of NowCast is as an estimate of what the daily average will be before the day is over. 

Notice that the NowCast was calculated from the monitor object including all days -- not just those included in the plot. This is because the NowCast calculation is based on previous data. For the best NowCast calculations, calculate from the full timeseries even if only a subset will be used for analysis or plotting. 

```{r nowcast}
# Calculate NowCast values
yrekaNowcast <- monitor_nowcast(yreka)

# Plot NowCast over rolling mean plot
monitorPlot_rollingMean(yrekaWeek, showLegend = FALSE)
monitorPlot_timeseries(yrekaNowcast, type = "l", add = TRUE, col = adjustcolor('royalblue', alpha.f = .7), 
                       tlim = c(20170121, 20170830), lwd = 2)
legend('topleft', legend = c("hourly PM2.5", "nowcast", "3-hour rolling mean"), pch = c(16, NA, NA), 
       lty = c(NA, 1, 1), col = c('gray80', adjustcolor('royalblue', alpha.f = .7), 'black'), lwd = c(NA, 2, 1))
addAQILegend()
```

While hourly timeseries plots provide detail about PM2.5 levels throughout the day, sometimes it is better to look at more aggregated data to eliminate distractions of sudden, short-lived spikes or drops. The dailyBarplot function will summarize each day by plotting the average PM2.5 value. 

```{r}
# Create a daily barplot
monitorPlot_dailyBarplot(yrekaWeek, labels_y_nudge = 12, labels_x_nudge = .2)
addAQILegend()
```

Another good way to summarize the data is by looking at the number of hours per day that the PM2.5 value was over a certain level. 

```{r daily_threshold}
# Calculate the number of hours over 'unhealthy'
unhealthyHours <- monitor_dailyThreshold(yrekaWeek, threshold =  "unhealthy")

# Plot unhealthy hours per day
barplot(unhealthyHours$data[,2], names.arg = strftime(unhealthyHours$data[,1], "%b %d"), cex.names = .8, ylim = c(0,25))
title("Hours where pm2.5 was unhealthy or worse", line = 2)

# Calculate the number of hours over 'very unhealthy'
veryUnhealthyHours <- monitor_dailyThreshold(yrekaWeek,threshold = "very unhealthy")

# Plot very unhealhty hours per day
barplot(veryUnhealthyHours$data[,2], names.arg = strftime(veryUnhealthyHours$data[,1], "%b %d"), cex.names = .8, ylim = c(0,25))
title("Hours where pm2.5 was very unhealthy or worse", line = 2)
```


Let's look at 2017 and see how air quality relates to major fires in the region. From the [Cal Fire report](http://cdfdata.fire.ca.gov/pub/cdf/images/incidentstatsevents_273.pdf) we can see that there were three fires in Siskiyou County which burned over 10,000 acres: The [Eclipse Complex](https://inciweb.nwcg.gov/incident/5511/), [Salmon August Complex](https://inciweb.nwcg.gov/incident/5501/), and [Orleans Complex](https://inciweb.nwcg.gov/incident/5430/). We can find their locations via (InciWeb)[https://inciweb.nwcg.gov/]. Here are their locations, with markers indicating the locations of temporary monitors:

```{r}
# Manually enter fire data
fireNames <- c("Eclipse Complex", "Salmon August Complex", "Orleans Complex")
fireLongitudes <- c(-123.493, -123.099, -123.647)
fireLatitudes <- c(41.792, 41.263, 41.566)
fireStartDate <- strptime(c("20170815", "20170813", "20170725"), format = "%Y%m%d")
fireEndDate <- strptime(c("20171010", "20171020", "20171012"), format = "%Y%m%d")
fireDF <- data.frame(name = fireNames, longitude =fireLongitudes, latitude = fireLatitudes, 
                     startDate = fireStartDate, endDate = fireEndDate)

# Map fires and monitors
monitorMap(sisk2017, ylim = c(40.8, 42.15), xlim = c(-124.4, -121), cex = 0)
addMarker(sisk2017$meta$longitude, sisk2017$meta$latitude, color = "blue")
addIcon('orangeFlame', fireDF$longitude, fireDF$latitude, expansion = .001)
text(fireLongitudes, fireLatitudes, fireNames, pos = 1)
```

Let's see if the timing of the fires lines up with spikes in PM2.5 concentrations. One way to summarize the data from the different monitors is by collapsing them into one. The PM2.5, latitude, and longitude values will be the means of the different monitors. 

```{r collapsed}
# Join the monitors into one monitor object with average PM2.5 values
collapsed2017 <- monitor_collapse(sisk2017, monitorID = "siskiyou_2017")

# Map monitors and collapsed monitor object
monitorMap(sisk2017, ylim = c(40.8, 42.15), xlim = c(-124.4, -121), cex = 0)
addMarker(sisk2017$meta$longitude, sisk2017$meta$latitude, expansion = .5, color = "blue")
addMarker(collapsed2017$meta$longitude, collapsed2017$meta$latitude, expansion = 1)
text(collapsed2017$meta$longitude, collapsed2017$meta$latitude, "collapsed \n monitor", pos=1, cex = .8)

# Plot timeseries of average values
par(mar = c(2, 4.5, 4, 2) + .1)
monitorPlot_timeseries(collapsed2017, tlim = c(20170701, 20171101), type = "l", ylim = c(-78, 400), col = "gray40", xlab = "")
title("Siskiyou County, CA 2017 fires and smoke")
abline(0,0,col = 'gray80')

# Add bars below the plot indicating the duration of each fire
rect(fireDF$startDate[1], -33, fireDF$endDate[1], -8, col = 'deepskyblue4', border = NA)
rect(fireDF$startDate[2], -63, fireDF$endDate[2], -38, col = 'darkolivegreen4', border = NA)
rect(fireDF$startDate[3], -93, fireDF$endDate[3], -68, col = 'darkorange3', border = NA)
text(fireDF$startDate, c(-20, -50, -80), fireDF$name, pos = 2, cex = .8)
```

We start seeing elevated PM2.5 levels shortly after the Orleans Complex is sparked. It's not until the middle of August, however, when two large fires start in the county that we really start seeing high PM2.5 concentrations. The PM2.5 concentrations drop to pre-fire levels around the middle of September and generally stay down, even as the fires continue burning well into October. 

