---
title: "Working with Generic Data"
author: "Spencer Pease"
date: "1/31/2019"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
options(knitr.table.format = "html")

library(PWFSLSmoke)
```

# Introduction

In this vignette, we take a look at how the `PWFSLSmoke` package works with
generic air quality data, given a data source and some basic configuration
details telling `PWFSLSmoke` how to interpret the data.


# Downloading Data


## Determining an Acceptable Data Format

First, we need to acquire some example data. AIRSIS provides access to raw data
in a `*.csv` format, one example of which can be downloaded [here][airsis-dl].
These AIRSIS files aren't yet in a format `generic_parseData()` accepts, so we
do some basic data cleaning to ensure each row corresponds to one observation.

```{r clean data, eval=FALSE}

## Note:
#  This section should only be run if a pre-cleaned version of the AIRSIS
#  example data does not already exist.

rawData <- readr::read_csv("localData/airsis_ebam_example-raw.csv")

fillCols <- list("Latitude", "Longitude", "Sys. Volts")

## NOTE: '!!!' is tidy unquoting for lists

cleanData <- rawData %>%
  tidyr::fill(!!!fillCols) %>%
  tidyr::fill(!!!fillCols, .direction = "up") %>%
  filter(!is.na(Type))

readr::write_csv(cleanData, "localData/airsis_ebam_example-clean.csv")

```

In general, `PWFSLSmoke` expects generic data to adhere to a few specific rules:

 * Each row corresponds to an observation at a single point in time
 * All of the required columns from the configuration list exist
 * Header rows only exist at the top of the file


## Get the Data

`generic_downloadData()` accepts a file path an argument. This path can be
either a local file path or a remote connection (such as `http://`, `https://`,
`ftp://`, or `ftps://). For this example, we will use a local file.

```{r download data}

filePath <- file.path("..", "localData", "airsis_ebam_example-clean.csv")
#fileString <- generic_downloadData(filePath)

```

The output of this function is the file, stored as a single string ready to be
parsed by `generic_parseData()`.


# Defining a Configuration List

```{r create configuration}

configurationList <- list(
  headerRows = 1,
  timezone = "UTC",
  columnTypes = "dcddc-dddddddddddc--dcc",
  requiredColumnNames = list(
    pm25 = "",
    datetime = ""
  ),
  extraColumnNames = list(
    monitorID = "",
    latitude = "",
    longitude = "",
    siteName = ""
  )
)

```


# Parsing Data





[airsis-dl]: http://usfs.airsis.com/vision/common/CSVExport.aspx?uid=1026&StartDate=2019-01-01&EndDate=2019-01-31
