---
title: "Working with Generic Data"
author: "Spencer Pease"
date: "1/31/2019"
output: html_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
options(knitr.table.format = "html")

library(PWFSLSmoke)
```

# Introduction

In this vignette, we take a look at how the `PWFSLSmoke` package works with
generic air quality data, given a data source and some basic configuration
details telling `PWFSLSmoke` how to interpret the data.


# Downloading Data


## Determining an Acceptable Data Format

First, we need to acquire some example data. AIRSIS provides access to raw data
in a `*.csv` format, one example of which can be downloaded [here][airsis-dl].
These AIRSIS files aren't yet in a format `generic_parseData()` accepts, so we
do some basic data cleaning to ensure each row corresponds to one observation.

```{r clean data, eval=FALSE}

## Note:
#  This section should only be run if a pre-cleaned version of the AIRSIS
#  example data does not already exist.

rawData <- readr::read_csv("localData/airsis_ebam_example-raw.csv")

fillCols <- list("Latitude", "Longitude", "Sys. Volts")

# The raw data consists of data records with no location information interspersed
# with 'GPS' records that have only location and voltage. This next step does
# the following:
#  1) replicate data associated with fillCols so that location information exists
#     in every row
#  2) remove any rows that are missing the "Type" field found in every data record

# NOTE:  '!!!' is tidy unquoting for lists
cleanData <- rawData %>%
  tidyr::fill(!!!fillCols, .direction = "down") %>%
  tidyr::fill(!!!fillCols, .direction = "up") %>%
  filter(!is.na(Type))

# Wow! Using tidyverse packages makes for very expressive code!

readr::write_csv(cleanData, "localData/airsis_ebam_example-clean.csv")

```

In general, `PWFSLSmoke` expects generic data to adhere to a few specific rules:

 * Each row corresponds to an observation at a single point in time
 * All of the required columns from the configuration list exist
 * Header rows only exist at the top of the file


## Get the Data

`generic_downloadData()` accepts a file path an argument. This path can be
either a local file path or a remote connection (such as `http://`, `https://`,
`ftp://`, or `ftps://`). For this example, we will use a local file.

```{r download data}

filePath <- filePath <- system.file(
  "extdata", "generic_data_example.csv",
  package = "PWFSLSmoke", mustWork = TRUE
)

fileString <- generic_downloadData(filePath)

```

The output of this function is the file, stored as a single string ready to be
parsed by `generic_parseData()`.


# Configuration List Details:

`configList` can be either an `R` `list` or a JSON file, but
both must follow the same nested key-value format:

```
 configList
  |_ column_names
  |   |_ datetime = [string],
  |   |_ pm25 = [string]
  |
  |_ station_meta
  |   |_ latitude = [numeric],
  |   |_ longitude = [numeric],
  |   |_ site_name = [optional | string]
  |
  |_ parsing_info
      |_ header_rows = [integer]
      |_ column_types = [string],
      |_ datetime_format = [string],
      |_ tz = [optional | string | default: "UTC"],
      |_ decimal_mark = [optional | string | default: "."],
      |_ grouping_mark = [optional | string | default: ","],
      |_ delimiter = [optional | string | default: ","],
      |_ encoding = [optional | string | default: "UTF-8"]
```

Required parameters are the top level keys of the configuration list.
Each parameter alters a specific aspect of parsing:

* "column_names" contains entries for mapping column names from the input
  file to `ws_monitor` columns names needed to built a `ws_monitor` object.

* "station_meta" contains entries for metadata about the station that
  generated the input file, which allows `PWFSLSmoke` to augment the
  `ws_monitor` object later in the pipeline. Some entries are required,
  others are optional.

* "parsing_info"

Extra keys in `columnNames` and `stationMeta` are currently ignored.


## Define a Configuration List

The configuration list can be read from a JSON file, or created as a list
directly.

```{r define configList}


# configListPath <- system.file(
#   "extdata", "generic_configList_example.json",
#   package = "PWFSLSmoke", mustWork = TRUE
# )
# configList <- jsonlite::fromJSON(configPath)

configList <- list(
  column_names = list(
    datetime = "Date/Time/GMT",
    pm25 = "ConcHr"
  ),
  station_meta = list(
    latitude = 35.75506,
    longitude = -118.4175,
    site_name = "1026 Kernville EBAM"
  ),
  parsing_info = list(
    header_rows = 5,
    column_types = "ccdddddddddddcdcc",
    datetime_format = "mdy T",
    tz = "UTC",
    delimiter = ","
  )
)

```


# Parse Data

```{r parse data}
parsedData <- generic_parseData(fileString, configList)

```



[airsis-dl]: http://usfs.airsis.com/vision/common/CSVExport.aspx?uid=1026&StartDate=2019-01-01&EndDate=2019-01-31
