---
title: "Test Vignette"
author: "Modified from a Mazama Science Original Document"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```
This is a sample vignette I put together during the second week of November, 2016.  I started with a vignette from the "PWFSLSmoke Tutorial" from the PWFSLSmoke/localVignettes folder.
```

## Background

Jon indicated last week that he wanted me to get familiar with documentation in R; specifically, R notebooks (via vignettes) and function documentation created using roxygen.  I spent a good part of last week working through the documentation for the monitor_* functions in the PWFSLSmoke package, and now I'm jumping into learning about vignettes.  It seemed the best way to learn would be to just create a sample vignette.  So, I copied an existing vignette and will be updating it below and testing all sorts of combinations so I can get a good handle on what is possible and how to do various tasks.

This is a **sample** vignette.  It serves the following purposes:

* Help me learn about vignettes
* Serve as a guide to the functionality available through vignettes

Let's get right to it, by creating a vignette that covers a few of the basics and then walks through some of the functions in the PWFSL R package.

## Table of Contents
* Basics
* 1: Load Data
* 2: Next things
* 3: More things after that.. 

## Basics

We'll go over a some of the basic functionality of R Markdowns.

### Imbedding and Running Code and Results

Let's play with a few different ways to work with and display `dim(iris)`. Note that for all code shown below, you'll want to substitute the ' characters with the diagonal tick mark (non-shift of "~" on the keyboard) which is the special character that RMarkdown recognizes as flagging R code.

Entering the following in R Markdown file
```
'''{r}
dim(iris)
'''
```
returns the following result (two lines):
```{r}
dim(iris)
```

Now let's try something a little different:
```
'''{r eval=FALSE}
dim(iris)
'''
```

This code yields, but this time no result is shown since eval=FALSE.
```{r eval=FALSE}
dim(iris)
```

And finally, here's the same code again, except this time we use echo=FALSE to hide the code itself, and instead only display the result.
```{r echo=FALSE}
dim(iris)
```

Note that I tried using the same code as above, while calling echo=FALSE and eval=FALSE, and R didn't like it at all, which makes sense, since why would you ever want to include code that you didn't want to show nor evaluate?  Anyway...

Now for some fun with in-line code.  Entering "Two plus two equals 'r 2 + 2'" (but with the smaller diagonal single quotes) yields the following -- note the part called by r in quotes was evaluated in R.

Two plus two equals `r 2 + 2`.

Note that the part in the small quotes above was evaluated, and the output was included within the text itself.

If you want to add a hyperlink to the vignette, you can [google](https://google.com) how to do it, or you can just look at this line of code in which case I have just demonstrated how to do this.

## Load Data

Before working with data, we have to load data into R.  We can do this as follows:

```{r eval = FALSE}
load("DNR_monitorList.RData")
load("DNR_rawList.RData")
combinedData <- monitor_combine(monitorList)
```

```
load("DNR_monitorList.RData")
load("DNR_rawList.RData")
combinedData <- monitor_combine(monitorList)
```


# Map the stations in our database, using the meta data in the ws_monitor object (created using monitor_combine above)
monitor_map(combinedData)


This package is designed to be used with [R](https://cran.r-project.org) (>= 3.1.0) and
[RStudio](http://rstudio.com) so make sure you have those installed first.

Users will want to install the **devtools** package to have access to latest versions
of some packages that are not yet available on CRAN.

The following packages should be installed with devtools by typing the following at the RStudio console:

```
devtools::install_github('rstudio/dygraphs')
devtools::install_github('rstudio/leaflet')
devtools::install_github('mazamascience/MazamaSpatialUtils', build_vignettes=TRUE)
devtools::install_github('mazamascience/PWFSLSmoke', build_vignettes=TRUE)
```

Any work with spatial data, *e.g.* assigning counties or watersheds, will require installation of required
spatial datasets. To get these datasets you should execute the following commands from a shell terminal:

```
mkdir ~/Data
cd ~/Data
curl -O http://smoke.airfire.org/RData/Spatial.tar.gz
tar -xzf Spatial.tar.gz
```


Best of luck exploring and understanding PM 2.5 values associated with wildfire smoke!