---
title: "TESTING_staticMap"
author: "Mazama Science"
date: "2019-04-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
suppressPackageStartupMessages({
  library(PWFSLSmoke)
  library(ggmap)
  library(raster)
})
```

# Trying out ggmap functions as a replacement for `esriMap_gteMap()`

First, what type of object does get_stamenmap() return?

```{r ggmap_object}
bbox <- c(left = -97.1268, bottom = 31.536245, right = -97.099334, top = 31.559652)
ggmap_obj <- ggmap::get_stamenmap(bbox, zoom = 13)
class(ggmap_obj)
typeof(ggmap_obj)
str(attributes(ggmap_obj))
length(ggmap_obj)
ggmap_obj[1:5]
```

It's a chartacter vector of RGB color values with extra attributes.

So what does the `esriMap_getMap()` download of a PNG file return?

```{r png_download}
response <- httr::GET("http://tile.stamen.com/terrain/13/1885/3338.png")
imageArray <- httr::content(response, type="image/png")
class(imageArray)
dim(imageArray)
rgbArray <- imageArray * 255
rgbArray[1:2,1:2,]
```

Can we convert the ggmap object into an `rgbArray`?

```{r ggmap_obj_to_rgbArray}
dim <- attr(ggmap_obj, "dim")
bb <- attr(ggmap_obj, "bb")
rgb_vec <- grDevices::col2rgb(ggmap_obj)
dim(rgb_vec)
red <- matrix(rgb_vec[1,], nrow = dim[2])
dim(red)
image(red)
```

It all worked and the image is upside down (as expected?)

Let's do the full png-to-mapRaster conversion seen in `esriMap_getMap()`

```{r ggmap_obj_to_mapRaster}
red <- matrix(rgb_vec[1,], nrow = dim[2])
green <- matrix(rgb_vec[2,], nrow = dim[2])
blue <- matrix(rgb_vec[3,], nrow = dim[2])
ggmap_rgbArray <- array(dim = c(dim[2], dim[1], 3))
ggmap_rgbArray[,,1] <- red
ggmap_rgbArray[,,2] <- green
ggmap_rgbArray[,,3] <- blue
ggmap_rgbArray[1:2,1:2,]
```

Can we now create a `mapRaster` from this?

Using more code from `esriMap_getMap()`:

```{r ggmap_obj_to_mapRaster}
# # Convert PNG into a Raster object
# mapRaster <- raster::brick(ncol=mapInfo$width,
#                            nrow=mapInfo$height,
#                            nl = 3)
# mapRaster <- raster::setValues(mapRaster, imageArray*255)
#  if (width == height) {mapRaster <- raster::t(mapRaster)} # rows and columns are confused when width = height
#
#  names(mapRaster) <- c("red", "green", "blue")
#
#  raster::extent(mapRaster) <- c(mapInfo$extent$xmin, mapInfo$extent$xmax, mapInfo$extent$ymin, mapInfo$extent$ymax)
#  raster::crs(mapRaster) <- sp::CRS(paste0("+init=epsg:",mapInfo$extent$spatialReference$latestWkid))

mapRaster <- raster::brick(ncol = dim[2],
                           nrow = dim[1],
                           nl = 3)
mapRaster <- raster::setValues(mapRaster, ggmap_rgbArray)
raster::extent(mapRaster) <- c(bb$ll.lon, bb$ur.lon, bb$ll.lat, bb$ur.lat)

# https://spatialreference.org/ref/sr-org/epsg3857-wgs84-web-mercator-auxiliary-sphere/
raster::crs(mapRaster) <- sp::CRS("+init=epsg:3857")
raster::extent(mapRaster)
raster::plotRGB(mapRaster)
```

Looking good! Let's try a larger area and see if we can put some points on it.

```{r washington}
bbox <- c(left = -125, bottom = 45, right = -119, top = 49)
ggmap_obj <- ggmap::get_stamenmap(bbox, zoom = 8)
dim <- attr(ggmap_obj, "dim")
bb <- attr(ggmap_obj, "bb")
rgb_vec <- grDevices::col2rgb(ggmap_obj)
red <- matrix(rgb_vec[1,], nrow = dim[2])
green <- matrix(rgb_vec[2,], nrow = dim[2])
blue <- matrix(rgb_vec[3,], nrow = dim[2])
ggmap_rgbArray <- array(dim = c(dim[2], dim[1], 3))
ggmap_rgbArray[,,1] <- red
ggmap_rgbArray[,,2] <- green
ggmap_rgbArray[,,3] <- blue
mapRaster <- raster::brick(ncol = dim[2],
                           nrow = dim[1],
                           nl = 3)
mapRaster <- raster::setValues(mapRaster, ggmap_rgbArray)
raster::extent(mapRaster) <- c(bb$ll.lon, bb$ur.lon, bb$ll.lat, bb$ur.lat)
raster::crs(mapRaster) <- sp::CRS("+init=epsg:3857")
raster::extent(mapRaster)
raster::plotRGB(mapRaster)

wa <- loadLatest() %>% monitor_subset(stateCode = "WA")
points(wa$meta$longitude, wa$meta$latitude, pch=16, col = "red")
```

Very close!
