---
title: "Data Model"
author: "Mazama Science"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.width=7, fig.height=5)
```

This vignette explores the `ws_monitor` data model used throughout the **PWFSLSmoke** package to store
and work with monitoring data.

The **PWFSLSmoke** package is designed to provide a compact, full-featured suite of utilities for
working with PM 2.5 data used to monitor wildfire smoke. A uniform data model
provides consistent data access across monitoring data available from different agencies.
The core data model in this package is defined by the `ws_monitor` object used to store data associated
with groups of individual monitors.

To work efficiently with the package it is important to understand the structure of this
data object and which functions operate on it. Package functions that begin with `monitor_`, 
expect objects of class `ws_monitor` as their first argument. (*Note that 'ws_' stands for 'wildfire smoke'.*)

## Monitoring Data

Monitoring data will typically be obtained from an agency charged with archiving data
acquired at monitoring sites. For wildifre smoke, the primary pollutant is PM 2.5 and the
sites archiving this data include [AirNow](https://www.airnow.gov), [WRCC](http://www.wrcc.dri.edu) and
[AIRSIS](https://app.airsis.com/USFS).

The data model for monitoring data consists of an **R** `list` with two dataframes: `data` and `meta`.

The `data` dataframe contains all hourly measurements organized with rows (the 'unlimited' dimension) as unique timesteps and
columns as unique monitoring sites. The very first column is always named `datetime` and contains the `POSIXct` datetime
in Coordinated Universal Time (UTC).

The `meta` dataframe contains all metadata associated with monitoring sites and is organized with
rows as unique sites and columns as site attributes. The following columns are guaranteed to exist
in the `meta` dataframe:

 * `monitorID` -- unique ID for each monitoring site
 * `longitude` -- decimal degrees East
 * `latitude` -- decimal degrees North
 * `elevation` -- meters above sea level
 * `timezone` -- Olson timezone
 * `countryCode` -- ISO 3166-1 alpha-2 code
 * `stateCode` -- ISO 3166-2 alpha-2 code

*(The [MazamaSpatialUtils](https://github.com/MazamaScience/MazamaSpatialUtils) package is used to assign timezones and state and country codes.)*

Starting with version 1.0 of the package, the following additional columns (mostly for internal use) will always exist:

 * `siteName` -- familiar name for a monitoring site
 * `agencyName` -- agency responsible for collecting the data
 * `countyName` -- county/province name
 * `msaName` -- US Census Bureau 'Metropolitan/Micropolitan Statistical Area'
 * `monitorType` -- broad instrument categories for E-Sampler, EBAM or BAM-1020
 * `monitorInstrument` -- specific instrument identifiers
 * `aqsID` -- AQS site identifier (often used as the `monitorID`)
 * `pwfslID` -- PWFSL site identifier (used as the `monitorID` for temporary monitors)
 * `telemetryAggregator` -- data provider for temporary monitors (e.g. 'wrcc' or 'usfs.airsis')
 * `telemetryUnitID` -- unique ID for each monitoring site used within the `telemetry_aggregator`

These columns of information are much more variable and, depending on the source of data, may include
many missing values.

It is important to note that the `monitorID` acts as a unique key that connects data with metadata.
The `monitorID` is used for column names in the `data` dataframe and for row names in the `meta` dataframe.
So the following will always be true:

    rownames(ws_monitor$meta) == ws_monitor$meta$monitorID
    colnames(ws_monitor$data) == c('datetime',ws_monitor$meta$monitorID)

**Example 1: Exploring `ws_monitor` Objects**

We will use the built-in "Northwest_Megafires" dataset and the `monitor_subset()` function to subset a `ws_monitor` object
which we can then explore.

```{r}
library(PWFSLSmoke)

# Get some airnow data for Washington state in the summer of 2015
N_M <- monitor_subset(Northwest_Megafires, tlim=c(20150801,20150831))
# To work with AirNow data directly, uncomment the next two lines
#N_M <- airnow_load(startdate=20150801, enddate=20150831)
#WA <- monitor_subset(airnow, stateCodes=c("WA"))
WA <- monitor_subset(N_M, stateCodes='WA')

# 'ws_monitor' objects can be identified by their class
class(WA)

# Examine the 'meta' dataframe
dim(WA$meta)
rownames(WA$meta)
colnames(WA$meta)

# Examine the 'data' dataframe
dim(WA$data)
colnames(WA$data)

# This should always be true
all(rownames(WA$meta) == colnames(WA$data[,-1]))
```
  
**Example 2: Manipulating `ws_monitor` Objects**

The **PWFSLSmoke** package has numerous functions that can work with `ws_monitor` objects, 
all of which begin with `monitor_`. If you need to do something that the package functions do not
provide you can manipulate `ws_monitor` objects directly as long as you retain the structure
of the data model.

The following code mixes use of package functions with direct manipulation of the `ws_monitor` object.

```{r}
# Use special knowledge of AirNow IDs to subset airnow data for Spokane county monitors
SpokaneCountyIDs <- N_M$meta$monitorID[stringr::str_detect(N_M$meta$monitorID, "^53063")]
Spokane <- monitor_subset(N_M, monitorIDs=SpokaneCountyIDs)

# Apply 3-hr rolling mean
Spokane_3hr <- monitor_rollingMean(Spokane, 3, align="center")

# 1) Replace data columns with their squares (exponentiation is not supplied by the package)
Spokane_3hr_squared <- Spokane_3hr
Spokane_3hr_squared$data[,-1] <- (Spokane_3hr$data[,-1])^2 # exclude the 'datetime' column

# NOTE:  Exponentiation is only used as an example. It does not generate a meaningful result.

# Create a daily averaged 'ws_monitor' object
Spokane_daily_3hr <- monitor_dailyStatistic(Spokane_3hr)

# 2) Check out the correlation between monitors (correlation is not supplied by the package)
data <- Spokane_daily_3hr$data[,-1] # exclude the 'datetime' column
cor(data, use='complete.obs')
```

----

This introduction to the `ws_monitor`  data model should be enough to get you started.
Lots more documentation and examples are available in the package documentation.

Best of luck exploring and understanding PM 2.5 values associated with wildfire smoke!
