---
title: "Poor Air Quality Levels In Denver On February 16, 2021"
author: "Mazama Science"
date: "2021-02-22"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
knitr::opts_chunk$set(error = TRUE)
```
```{r libraries, echo = FALSE, eval=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
library(PWFSLSmoke)
library(magrittr)
```
On Tuesday morning (February 16, 2021), Meteorologist Ashton Altieri 
[predicted](https://denver.cbslocal.com/2021/02/16/denver-weather-air-quality-quickly-worsen-tuesday-more-snow-possible/) 
worrying air quality levels in the Denver metro area and along the entire Front Range through the afternoon. 
The Colorado Department of Public Health and Environment encouraged to limit 
driving non-EV vehicles in Denver, Boulder and Fort Collins to avoid worsening 
the air pollution levels.

In Colorado, poor air quality levels during winter are due to *temperature inversions* - 
"weather events in which cold, polluted air near the ground is unable to 
escape into the atmosphere and disperse because of a layer of hotter air above, 
which creates a trapping effect. In the winter, weakened sunlight and a frozen 
ground combine to create temperature inversions, sometimes lasting days at a 
time" ([Air quality in Denver](https://www.iqair.com/us/usa/colorado/denver)). 

Let's have a look at the PM2.5 levels around Tuesday by starting from 
a map of the affected area.

```{r, monitor IDs, echo=FALSE, include=FALSE}
monitorIDs = c("080310002_01", "080131001_01", "080130003_01", "081230006_01", 
               "080690009_01")
```

```{r, airnow_loadLatest and subset, echo=FALSE}
CO = airnow_loadLatest(parameter = "PM2.5") %>%
  monitor_subset(monitorIDs = monitorIDs, tlim = c(20210215,20210218))
```

```{r, leaflet, echo=FALSE}
monitor_leaflet(CO, legendTitle = "Denver Metro Area",
               maptype = "Esri.WorldTopoMap")
```

<br><br>

Indeed, the map shows that the main cities along the Front Range experienced 
poor air quality indexes (AQI) with PM2.5 reaching USG (unhealthy for sensitive group) and unhealthy levels.

We can also explore further the AQI levels using different types of plots:

### Dynamic Graph
```{r, monitor_dygraph, echo=FALSE}
monitor_dygraph(CO, title = "Front Range AQIs - Feb 15 - 18, 2021", 
                showLegend = TRUE)

```

<br>

### Timeseries Plot avaraging the AQI levels of more cities 
```{r, monitor_timeseriesPlot, echo=FALSE}
monitor_timeseriesPlot(CO, style = "gnats")
addAQIStackedBar()
addAQILines()
title("Front Range AQIs - Feb 15 - 18, 2021")
```

<br>

### Timeseries Plot showing the AQI levels of each city individually
```{r, Denver - CAMP - 2105 Broadway. Monitor ID: 080310002_01, echo=FALSE}
monitorID_D = "080310002_01"
Denver = airnow_loadLatest(parameter = "PM2.5") %>%
  monitor_subset(monitorIDs = monitorID_D, tlim = c(20210215,20210218))
monitor_timeseriesPlot(Denver, style = "gnats")
addAQIStackedBar()
addAQILines()
title("Denver AQI - Feb 15 - 18, 2021")
```

<br>
```{r, Boulder - CU - 2102 Athens St. Monitor ID: 080131001_01, echo=FALSE}
monitorID_D = "080131001_01"
Denver = airnow_loadLatest(parameter = "PM2.5") %>%
  monitor_subset(monitorIDs = monitorID_D, tlim = c(20210215,20210218))
monitor_timeseriesPlot(Denver, style = "gnats")
addAQIStackedBar()
addAQILines()
title("Boulder AQI - Feb 15 - 18, 2021")
```

<br>
```{r, Longmont. Monitor ID: 080130003_01, echo=FALSE}
monitorID_D = "080130003_01"
Denver = airnow_loadLatest(parameter = "PM2.5") %>%
  monitor_subset(monitorIDs = monitorID_D, tlim = c(20210215,20210218))
monitor_timeseriesPlot(Denver, style = "gnats")
addAQIStackedBar()
addAQILines()
title("Longmont AQI - Feb 15 - 18, 2021")
```

<br>
```{r, Greeley Hospital. Monitor ID: 081230006_01, echo=FALSE}
monitorID_D = "081230006_01"
Denver = airnow_loadLatest(parameter = "PM2.5") %>%
  monitor_subset(monitorIDs = monitorID_D, tlim = c(20210215,20210218))
monitor_timeseriesPlot(Denver, style = "gnats")
addAQIStackedBar()
addAQILines()
title("Greeley AQI - Feb 15 - 18, 2021")
```

<br>
```{r, Ft. Collins - CSU Facilities. Monitor ID: 080690009_01, echo=FALSE}
monitorID_D = "080690009_01"
Denver = airnow_loadLatest(parameter = "PM2.5") %>%
  monitor_subset(monitorIDs = monitorID_D, tlim = c(20210215,20210218))
monitor_timeseriesPlot(Denver, style = "gnats")
addAQIStackedBar()
addAQILines()
title("Ft. Collins AQI - Feb 15 - 18, 2021")
```

<br>
As shown in the plots, Greeley experienced the highest PM2.5 levels (~70 µg/m3) 
while Boulder was the least affected (~40 µg/m3).

The air quality monitors used for this analysis were selected based on their geographic location by using 
the 
[AirFire Monitoring PM2.5 tool](https://tools.airfire.org/monitoring/v4/#!/?category=PM2.5_nowcast&centerlat=42&centerlon=-95&zoom=4). 


```{r, list of monitors, echo=FALSE, include=FALSE}
# * Denver - CAMP - 2105 Broadway. Monitor ID: 080310002_01
# 
# * Boulder - CU - 2102 Athens St. Monitor ID: 080131001_01
# 
# * Longmont. Monitor ID: 080130003_01
# 
# * Greeley Hospital. Monitor ID: 081230006_01
# 
# * Ft. Collins - CSU Facilities. Monitor ID: 080690009_01
```


