---
title: "Example: Denver Air Quality Forecast"
author: "Mazama Science"
date: "2021-02-23"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example: Denver Air Quality Forecast}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, echo = FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
knitr::opts_chunk$set(error = TRUE)
```

On Tuesday morning (February 16, 2021), Meteorologist Ashton Altieri 
[predicted](https://denver.cbslocal.com/2021/02/16/denver-weather-air-quality-quickly-worsen-tuesday-more-snow-possible/) 
worrying air quality levels in the Denver metro area and along the entire Front 
Range through the afternoon. The Colorado Department of Public Health and 
Environment encouraged people to limit driving non-EV vehicles in Denver, Boulder
and Fort Collins to avoid worsening the air pollution levels.

In Colorado, poor air quality levels during winter are due to *temperature inversions* - 
"weather events in which cold, polluted air near the ground is unable to 
escape into the atmosphere and disperse because of a layer of hotter air above, 
which creates a trapping effect. In the winter, weakened sunlight and a frozen 
ground combine to create temperature inversions, sometimes lasting days at a 
time" ([Air quality in Denver](https://www.iqair.com/us/usa/colorado/denver)). 

Let's use the PWFSLSmoke R package to do our own investigation of the PM2.5 
levels around that date to see if the forecast panned out.

## PWFSLSmoke R package

We will begin by loading the **dplyr** and **PWFSLSmoke** packages and loading
the latest data so we can find some monitors of interest between Denver and
Fort Collins:

```{r load_latest, echo = TRUE, warning = FALSE, message = FALSE}
library(dplyr)
library(PWFSLSmoke)

# Load, subset and plot recent PM2.5 data
monitor_loadLatest() %>%
  monitor_subset(stateCodes = "CO") %>%
  monitor_leaflet()
```


Zooming in and clicking on individual monitoring sites, we can create a collection
of monitorIDs for which we will build a _ws_monitor_ object for further inspection

```{r custom_monitor_object, echo = TRUE, warning = FALSE, message = FALSE}
# Vector of monitorIDs
monitorIDs <- c("080310002_01", "080131001_01", "080130003_01", "081230006_01", 
               "080690009_01")

# Use MazamaCoreUtils::dateRange() to ensure we get POSIXct times in the local timezone
dateRange <- MazamaCoreUtils::dateRange(20210215, 20210218, timezone = "America/Denver")

# Create a 'ws_monitor' object with monitors of interest
my_monitors <- 
  monitor_load(dateRange[1], dateRange[2]) %>%
  monitor_subset(monitorIDs = monitorIDs)

# Another leaflet map showing max AQI values
monitor_leaflet(my_monitors)
```


Indeed, the map shows that the main cities along the Front Range experienced 
poor air quality indexes (AQI) with PM2.5 reaching USG (unhealthy for sensitive 
groups) and Unhealthy levels.

We can explore PM2.5 levels in detail with a "dygraph" interactive timeseries
plot:

```{r, monitor_dygraph, echo = TRUE, warning = FALSE, message = FALSE}
monitor_dygraph(
  my_monitors, 
  title = "Front Range AQIs - Feb 15 - 18, 2021", 
  showLegend = TRUE
)
```

The **PWFSLSmoke** package also provides a baseplot timeseries plot that can
be used to show all the data points in a publication-ready graphic:

```{r, monitor_timeseriesPlot, echo = TRUE, warning = FALSE, message = FALSE}
monitor_timeseriesPlot(my_monitors, style = "gnats")
addAQIStackedBar()
addAQILines()
addAQILegend()
title("Front Range PM2.5 values -- Feb 15 - 18, 2021")
```

The `monitor_timeseriesPlot()` function allows a variety of visualizations to
be created:

```{r, monitor_timeseriesPlot_2, echo = TRUE, warning = FALSE, message = FALSE}
monitor_timeseriesPlot(my_monitors, type = 'l')
monitor_timeseriesPlot(my_monitors, style = 'aqidots', pch=16, cex = 1, add = TRUE)
addAQIStackedBar()
addAQILines()
addAQILegend()
title("Front Range PM2.5 values -- Feb 15 - 18, 2021")
```

## AirMonitorPlots R package

For those with greater experience with R, the
[AirMonitorPlots](https://mazamascience.github.io/AirMonitorPlots/index.html) 
package provides additional, **ggplot2** based graphics for _ws_monitor_ objects 
from the **PWFSLSmoke** package. **AirMonitorPlots** has not been released to 
CRAN but can be installed with:

```
devtools::install_github("MazamaScience/AirMonitorPlots")
```

The following example demonstrates what is possible with **AirMonitorPlots**: 

```{r ggplot_pm25Timeseries, fig.height = 7, echo = TRUE, warning = FALSE, message = FALSE}
# After looking at my_monitors$meta$siteName, help ggplot out by assigning 
# short labels to be used the facet label
my_monitors$meta$shortName <- c(
  "Denver", "Boulder", "Longmont", "Greeley", "Ft. Collins"
)

# Load AirMonitorPlots
library(AirMonitorPlots)

# Fancy plot with AQI bars
gg <- 
  ggplot_pm25Timeseries(
    my_monitors,
    dateRange[1],
    dateRange[2],
    timezone = "America/Denver"
  ) +
  ggtitle("Front Range PM2.5 values -- Feb 15 - 18, 2021") +
  stat_AQCategory(color = NA) + 
  facet_grid(rows = vars(shortName)) 

print(gg)
```


Greeley experienced the highest PM2.5 levels (~70 µg/m3) 
while Boulder was the least affected (~40 µg/m3). All in all, the prediction of
worsening air quality during the day on Tuesday was largely realized as air 
quality reached USG levels in all communities and Unhealthy in Greeley.

-----

_NOTE:_  Up-to-the hour PM2.5 data from official monitors as well as low-cost 
sensors is available at https://fire.airnow.gov.
