---
title: "Loading Monitoring Data"
author: "Mazama Science"
date: "2021-02-26"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Loading Monitoring Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)
knitr::opts_chunk$set(error = TRUE)
```
```{r libraries, echo = FALSE, eval=TRUE, warning=FALSE, message=FALSE}
library(dplyr)
library(PWFSLSmoke)
library(magrittr)
```

This article describes pre-generated `ws_monitor` objects available from a 
USFS maintained archive and how to load them with various `monitor_load~()`
functions.

## Pre-Generated Data Files

The USFS Pacific Wildland Fire Sciences Lab [AirFire](https://www.airfire.org) 
team works to model wildland fire emissions and maintains an up-to-date database
of data from monitors using data from regulatory-type monitors
([FRM or FEM](https://www.epa.gov/amtic/air-monitoring-methods-criteria-pollutants)).

Instruments include _permanent_ monitors that form part of the 
[AirNow](https://www.airnow.gov) network as well as _temporary_ monitors that
are deployed during wildfire events. Quality controlled data from permanent 
monitors comes directly from AirNow. Data from temporary monitors is provided in
raw form from either [AIRSIS](https://app.airsis.com/USFS) or the Western 
Regional Climate Center [WRCC](http://www.wrcc.dri.edu) and is aggregated and
quality controlled by AirFire in real-time.

As of February, 2021, pre-generated `ws_monitor` objects containing this data
are publicly available at https://haze.airfire.org/monitoring/.

## Four types of archive files — hourly, daily, monthly, and annual 

The data archive is maintained by AirFire and, for each data provider, the 
archive offers datasets that are updated in real-time and on hourly-, daily-, 
and monthly schedules The following types of files are available: 

**Hourly files** -- These "real-time" files are updated multiple times an hour 
(~ every two minutes) and contain data for the previous 10 days. They are used
to generate graphics for the [Monitoring v4 site](https://tools.airfire.org/monitoring/v4).

**Daily files** -- Daily files are updated once a day, shortly after midnight and
contain data for the previous 45 days. The time span is long enough to contain 
data for a previous month until 10-days into the next month at which point
AirNow stops _backfilling_ data that straggles in due to power outages, _etc._

**Monthly files** -- Monthly files are updated within a few days after the 10'th 
of the following month after data updates have ceased. For instance, a July file
will be updated in mid-August. 

**Annual files** -- Annual files are updated mid-month, after the monthly file
update, by appending the just-updated monthly file (covering the previous month)
to the growing annual file. Thus, an annual will will always cover the period 
from the beginning of the year to the end of the last month. 
 
*NOTE*: The date and time of annual and monthly updates are subject to change. 
Exploring the [data archive](https://haze.airfire.org/monitoring/) with a web 
browser, the exact date and time is seen next to file of interest, under the column 
`Last modified`. For instance, in the 
[WRCC 2017 archive](https://haze.airfire.org/monitoring/WRCC/RData/2017/) you can 
see that the file `wrcc_PM2.5_2021.RData` was generated on February 14, 2021 at 
01:24 am. 

## Archive Structure

The AirFire maintained data archive uses a consistent and predictable structure:

```
├── AIRSIS
│   └── RData
│       ├── 2004
│       ├── ...
│       └── 2021
├── AirNow
│   └── RData
│       ├── 2016
│       ├── ...
│       ├── 2020
│       │   ├── 01
│       │   ├── ...
│       │   └── 12
│       └── 2021
│           ├── 01
│           └── 02
├── EPA
│   └── RData
│       ├── 1998
│       ├── ...
│       └── 2017
├── WRCC
│   └── RData
│       ├── 2010
│       ├── ...
│       └── 2021
└── latest
    └── RData
```

**Top level** -- There is a dedicated folder for each agency 
(i.e [AIRSIS/](https://haze.airfire.org/monitoring/AIRSIS/), 
[AirNow/](https://haze.airfire.org/monitoring/AirNow/),
[WRCC/](https://haze.airfire.org/monitoring/WRCC/)). Additionally, at this level,
latest data (hourly and daily updates) for all agencies is grouped together and 
can be quickly accessed from the 
[latest/RData/](https://haze.airfire.org/monitoring/latest/RData/) subdirectory.

**Second level** -- Each agency has the `RData/` folder (_e.g._ A`irNow/RData/`)
where all R binary files are stored. Other data types like `csv/` and `geojson/` 
may also exist.

**Third level** -- Each data provider has annual folders (_e.g._ `AirNow/RData/2021/`) 
and a `latest/` folder (_e.g._ `AirNow/RData/latest/`). The latter contains hourly
(e.g. airnow_PM2.5_latest10.RData) and daily (e.g. `airnow_PM2.5_latest45.RData`) 
updates. (This a just copy of the same data found in `latest/RData/`.)

**Fourth level** -- Each data provider has annual datasets 
(_e.g._ `airnow_PM2.5_2021.RData`). Additionally, because of the size of the files,
AirNow has monthly folders (_e.g._ `AirNow/RData/2021/01/`), and 
therefore a **fifth level** containing monthly datasets 
(_e.g._ `airNow_PM2.5_2021_01.RData`). 

*NOTE*: No annual data files exist for AirNow prior to 2016.

## Direcly loading datasets

With knowledge of the structure of the data archives, it is possible to load 
datasets directly from the data archive with base R functions by using 
`get(load(url(…)))`. For instance, if you want to load the latest data 
(hourly updates) from AirNow, you can just:

```{r, example for loading data, eval=FALSE}
AirNow_hourly <- get(load(url("https://haze.airfire.org/monitoring/latest/RData/airnow_PM2.5_latest10.RData")))
```


## Loading functions

For each agency (AirNow, AIRSIS, WRCC), the 
[PWFSLSmoke package](http://mazamascience.github.io/PWFSLSmoke/reference/index.html) 
offers three loading functions: 

`~_loadLatest()` - for hourly updates covering the most recent 10 days

```{r, example loadLatest , eval=FALSE}
airnow_loadLatest(
parameter = 'PM2.5',
baseUrl = "https://haze.airfire.org/monitoring/latest/RData",
dataDir = NULL) 
```

~__loadDaily()` - for daily updates covering the most recent 45 days

```{r, example loadDaily, eval=FALSE}
airnow_loadDaily(
parameter = 'PM2.5',
baseUrl = "https://haze.airfire.org/monitoring/latest/RData",
dataDir = NULL) 
```

`~_loadAnnual()` - for data extended more than 45 days into the past 

```{r, example loadAnnual, eval=FALSE}
airnow_loadAnnual(
year = NULL,
parameter = 'PM2.5',
baseUrl = "https://haze.airfire.org/monitoring",
dataDir = NULL) 
```

These functions are typically used with default settings for `parameter`,
`baseUrl` and `dataDir`:

```
airnow_latest <- airnow_loadLatest()
wrcc_daily <- wrcc_lodDaily)
```

_ NOTE:_ If `dataDir` is specified, data will be preferentially loaded from a 
local directory rather than from `baseUrl`. 

## `epa_load()`

The `epa_loadd()` and `epa_loadAnnual()` functions have an additional
`parameterCode` value which must be specified. EPA parameter codes include:

* `88101` – PM2.5 FRM/FEM Mass (begins in 2008)
* `88502` – PM2.5 non FRM/FEM Mass (begins in 1998)

## `monitor_load()`

```{r, example load , eval=FALSE}
monitor_load(
  startdate = NULL,
  enddate = NULL,
  monitorIDs = NULL,
  parameter = "PM2.5",
  baseUrl = "https://haze.airfire.org/monitoring",
  dataDir = NULL,
  aqsPreference = "airnow"
)
```

The preferred function for general data loading is 
[monitor_load()](http://mazamascience.github.io/PWFSLSmoke/reference/monitor_load.html), 
which loads all available monitoring data for a given time range. Data from 
AirNow, AIRSIS and WRCC are combined into a single *ws_monitor* object. 
Archival datasets are joined with 'daily' and 'latest' datasets as needed to 
satisfy the requested date range.

*NOTE*: Because `monitor_load()` performs a lot of validation tasks as it joins
annual, monthly and recent data, it is recommended that you load no more than ~3 months 
of consecutive data using this function. Longer timeseries can be obtained using
the `_loadAnnual()` functions.

*NOTE*: the **PWFSLSmoke** package does not _yet_ create time series cross year boundaries 
because “joining” timeseries together can take a long time. This capability will
be enabled at some point in the future.

## Examples using PWFSLSmoke loading functions

### Timeseries plot using the latest data for Washington State 

```{r, loadLatest WA, warning=FALSE}
# Recipe to get latest AirNow data for WA
airnow_wa <-
  airnow_loadLatest() %>%
  monitor_subset(stateCodes= 'WA')

# Timeseries plot
monitor_timeseriesPlot(airnow_wa, style = 'gnats')
addAQIStackedBar()
addAQILines()
title("Washington State Latest AirNow Data")
```

### Timeseries plot using 2015 annual data for the Pacific Northwest. 

```{r, OR_ID_WA airsis_loadAnnual(2015), warning = FALSE}
# EPA FRM/FEM data
epa <-
  epa_loadAnnual(2015, parameterCode = "88101") %>%
  monitor_subset(
    stateCodes = c("OR", "ID", "WA"),
    tlim = c(20150701, 20151001)
  )

# AIRSIS temporary monitors
airsis <-
  airsis_loadAnnual(2015) %>%
  monitor_subset(
    stateCodes = c("OR", "ID", "WA"),
    tlim = c(20150701, 20151001)
  )

# Combine EPA and AIRSIS data into a single ws_monitor object
pnw <-
  monitor_combine(list(epa, airsis)) %>%
  monitor_nowcast()

# Timeseries plot using NowCast smoothing
monitor_timeseriesPlot(pnw, style = 'gnats')
addAQIStackedBar()
addAQILines()
title("Pacific Northwest 2015 Megafires -- NowCast Smoothed")
```

----

_Best of luck gathering data for your research!_
