---
title: "PWFSLSmoke's Data Model"
author: "Mazama Science"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The **PWFSLSmoke** package is designed to provide a compact, full-featured suite of utilities for
working with PM 2.5 data used to monitor wildfire smoke. Uniform data models
provides consistent data access across monitoring data available from different agencies.

The core data model in this package is the `ws_monitor` object used to store data associated
with groups of individual monitors.

To work efficiently with the package it is important to understand the structure of this
data object and which functions operate on it. Package functions that begin with `monitor_`, 
expect objects of class `ws_monitor` as their first argument. (*Note that 'ws_' stands for 'wildfire smoke'.*)

## Monitoring Data

Monitoring data will typically be obtained from an agency charged with archiving data
acquired at monitoring sites. For wildifre smoke, the primary pollutant is PM 2.5 and the
sites archiving this data include [AirNow](http://www.airnow.gov), [WRCC](http://www.wrcc.dri.edu) and
[AIRSIS](http://app.airsis.com/USFS/). For this tutorial we will be getting out data from WRCC. 

We can begin by creating a `ws_monitor` and looking at how it's made. Once we get a general look at a `ws_monitor` object we will start do some examples to get more in depth knowledge of them. Try out the following code to get started:

```
library(PWFSLSmoke)

monitor_data <- wrcc_load('20150801', '20150810')
```

This is going to load wrcc monitor data from August 1-10 from 2015. We can look at what `monitor_data` is by exploring it's class:

```
class(monitor_data)
> [1] "ws_monitor" "list"

class(monitor_data$data)
> [1] "data.frame"

class(monitor_data$meta)
[1] "data.frame"
```

We can see that `monitor_data` is a list containing two dataframes: `data` and `meta`.

The `data` dataframe contains all hourly measurements organized with rows (the 'unlimited' dimension) as unique timestamps and
columns as unique monitors. The very first column is always named `datetime` and contains the `POSIXct` datetime
in Universal Time Coordinates (UTC = GMT).

The `meta` dataframe contains all metadata associated with monitoring sites and is organized with
rows as unique sites and columns as site attributes. The following columns are guaranteed to exist
in the `meta` dataframe:

 * `monitorID` -- unique ID for each monitoring site (*i.e.* each instrument deployment)
 * `longitude` -- decimal degrees East
 * `latitude` -- decimal degrees North
 * `elevation` -- meters above sea level
 * `timezone` -- Olson timezone
 * `countryCode` -- ISO 3166-1 alpha-2 code
 * `stateCode` -- ISO 3166-2 alpha-2 code
 
*(The [MazamaSpatialUtils](https://github.com/MazamaScience/MazamaSpatialUtils) package is used to assign timezones and state and country codes.)*
 
Let's look and see what columns out `monitor_data` has:
```
names(monitor_data$meta)
> [1] "AQSID"          "siteCode"       "siteName"       "status"        
> [5] "agencyID"       "agencyName"     "EPARegion"      "latitude"      
> [9] "longitude"      "elevation"      "timezone"       "GMTOffsetHours"
>[13] "countryCode"    "FIPSMSACode"    "MSAName"        "FIPSStateCode" 
>[17] "stateCode"      "GNISCountyCode" "countyName"     "monitorID"  
```

As we can see, we have all the guaranteed columns and then some!

It is important to note that the `monitorID` acts as a unique key that connects data with metadata.
The `monitorID` is used for column names in the `data` dataframe and for row names in the `meta` dataframe.
So the following will always be true for any `ws_monitor`:

    rownames(ws_monitor$meta) = ws_monitor$meta$monitorID
    colnames(ws_monitor$data) = c('datetime',ws_monitor$meta$monitorID)
 
**Example 1: Exploring `ws_monitor` Objects**

We will now create and manipulate a new `ws_monitor` object with data from airnow: 

```{r}
library(PWFSLSmoke)

# Set the location of the local data archive
setSmokeDataDir('~/Data/Smoke')

# Get some airnow data for Washington
airnow <- airnow_load(startdate=20150801, enddate=20150831)
WA <- monitor_subset(airnow, stateCodes='WA')

# 'ws_monitor' objects can be identified by their class
class(WA)

# Examine the 'meta' dataframe
dim(WA$meta)
rownames(WA$meta)
colnames(WA$meta)

# Examine the 'data' dataframe
dim(WA$data)
colnames(WA$data)

# This should always be true
all(rownames(WA$meta) == colnames(WA$data[,-1]))
```
  
**Example 2: Manipulating `ws_monitor` Objects**

The **PWFSLSmoke** package has numerous functions that can work with `ws_monitor` objects, 
all of which begin with `monitor_`. If you need to do something that the package functions do not
provide you can manipulate `ws_monitor` objects directly as long as you retain the structure
of the data model.

The following code mixes use of package functions with direct manipulation of the `ws_monitor` object.

```{r}
# Use special knowledge of AirNow IDs to subset airnow data for Spokane county monitors
SpokaneCountyIDs <- airnow$meta$monitorID[stringr::str_detect(airnow$meta$monitorID, "^53063")]
Spokane <- monitor_subset(airnow, monitorIDs=SpokaneCountyIDs)

# Apply 3-hr rolling mean
Spokane_3hr <- monitor_rollingMean(Spokane, 3, align="center")

# Replace data columns with their squared values (exponentiation is not supplied by the package)
Spokane_3hr$data[,-1] <- (Spokane_3hr$data[,-1])^2

# NOTE:  Exponentiation is only used as an example. It does not generate a meaningful result.

# Create a daily averaged 'ws_monitor' object
Spokane_daily_3hr <- monitor_dailyStatistic(Spokane_3hr)

# Check out the correlation between monitors (correlation is not supplied by the package)
data <- Spokane_daily_3hr$data[,-1] # omit the 'datetime' column
cor(data, use='complete.obs')
```

----

This introduction to the `ws_monitor`  data model should be enough to get you started.
There are additional tutorials that will go more in depth of how to obtain, manipulate and visualilze `ws_monitor` that can be used to learn more about the  PWFSLSmoke package as well as also serve as a reference. 


----
 
 
next >
[Obtaining Data](Obtaining Data)